Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-08-28 paper-amia-summit-2018/amia_summit_2018.tex
update data collection, evaluation metrics and conclusion

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 37 motivational interviews, which include a total of 11,353 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 1072 observed sequences, 880 were positive and 192 were negative. Alternating communication sequences are sub-sequences of a natural communication sequence, in which two consecutive utterances cannot belong to the same speaker (i.e. a counselor behavior code has to be followed by an adolescent behavior code and vice versa). For example, 6 alternating sequences (\{C1, A1, C2, A3\}, \{C1, A1, C3, A3\}, \{C1, A1, C4, A3\}, \{C1, A2, C2, A3\}, \{C1, A2, C3, A3\} and \{C1, A2, C4, A3\}) can be created from a natural sequence \{C1, A1, A2, C2, C3, C4, A3\}, where adolescent and counselor codes are represented by A and C, respectively. A total of 18753 alternating sequences have been observed in experimental dataset, out of which 18139 were successful and 614 were unsuccessful ones. For each type of sequence (natural and alternating), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences. Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. \\

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
Successful sequences & 880 & 82.09\% & 5.83 \\\hline
Unsuccessful sequences & 880 & 82.09\% & 5.83 \\\hline
  \end{tabular}
\end{table}

Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. \\

@@

\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, T-AMB, CCT, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/delta.png}
    \caption{\textbf{F-Measure of the proposed Markov model-based method by varying the value of $\delta$}.}
    \label{fig:delta}
\end{figure}

@@

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)).

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

@@

\section*{Conclusion}
In this paper, we proposed a method based on Markov Chain and Hidden Markov Model for predicting the success of motivational interviews. We found out that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success. Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the  interviews without resource intensive manual qualitative analysis of interview transcripts. Our proposed method can also help identify specific patterns that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice. This work also has broad implications for qualitative public health research by providing a formal theoretically-grounded computational mechanism to facilitate the development of effective behavioral interventions.

>>>>

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 3427 were positive and 742 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. \\

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
Successful sequences & 3427 & 82.20\% & 10.05 \\\hline
Unsuccessful sequences & 742 & 17.80\% & 9.89 \\\hline
  \end{tabular}
\end{table}

As can be seen in Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels. In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set. Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm. Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. \\

@@

\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:

@@

@@

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folda cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). However, Accuracy is simply computed as the ratio of correctly classified sequences and total number of sequences.

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

@@

\section*{Conclusion}
In this paper, we proposed Recurrent Neural Networks with two baseline methods Markov Chain and Hidden Markov Model for predicting the success of motivational interviews. We found out that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success. Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the interviews without resource intensive manual qualitative analysis of interview transcripts. Our proposed method can also help to identify specific patterns that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice. This work also has broad implications for qualitative public health research by providing a formal theoretically-grounded computational mechanism to facilitate the development of effective behavioral interventions.
